### 2025-03-14 paper/quarto_ecml/paper.tex
Shorten to slightly less than 16 pages by restructuring

\textbf{Specifically, we propose counterfactual training (CT)}: a novel
training regime that aligns learned representations with plausible
explanations compliant with user requirements. The remainder of this
paper is structured as follows. Section~\ref{sec-lit} presents related
work, focusing on the link between adversarial examples and
counterfactual explanations. Then follow our main contributions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In Section~\ref{sec-method}, we introduce our methodological framework
  and show theoretically that it can be used to enforce global
  actionability constraints.\\
\item
  In Section~\ref{sec-experiments}, through extensive experiments we
  demonstrate that CT substantially improves explainability without
  sacrificing predictive performance.
\end{enumerate}

We discuss the challenges in Section~\ref{sec-discussion} and conclude
in Section~\ref{sec-conclusion} that CT is a promising approach towards
making opaque models more trustworthy.

@@

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation.
\cite{wu2021polyjuice} propose \emph{Polyjuice}, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that the augmentation of training data through
\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
tasks. \cite{balashankar2023improving} similarly use \emph{Polyjuice} to
augment NLP datasets through diverse counterfactuals and show that
classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure. First, the
authors identify training samples that are subject to high predictive
uncertainty. Second, they generate counterfactual explanations for those
samples. Finally, they fine-tune the given language model on the
augmented dataset that includes the generated counterfactuals.

@@

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that guarantee
individual recourse to some positive target class with high probability.
Their approach builds on adversarial training by explicitly inducing
susceptibility to targeted adversarial examples for the positive class.
Additionally, the proposed method allows for imposing a set of
actionability constraints ex-ante. For example, users can specify that
certain features (e.g., \emph{age}, \emph{gender}) are immutable.
Secondly, \cite{guo2023counternet} are the first to propose an
end-to-end training pipeline that includes counterfactual explanations
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to \cite{guo2023counternet},
we impose no restrictions on the neural network architecture at all.

@@

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

@@

From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this simply in terms ignoring
implausibility costs with respect to immutable features, which
effectively forces the model to instead seek plausibility with respect
to the remaining features. This in turn results in lower overall
sensitivity to immutable features, which we demonstrate empirically for
different classifiers in Section~\ref{sec-experiments}. Under certain
conditions, this results holds theoretically:\footnote{For the proof,
  see the supplementary appendix.}

@@

It is worth highlighting that \mbox{Prp. \ref{prp-mtblty}} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxies that remain
unprotected. We address this in Section~\ref{sec-discussion}.

@@

We have simulated this scenario using synthetic data with two
independent features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
outcomes for different training procedures using the same model
architecture each time (a linear classifier). In each case, we show the
decision boundary (in green) and the training data colored according to
their ground-truth label: orange points belong to the target class,
\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
indicate counterfactuals in the target class generated at test time
using generic gradient descent until convergence.

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they do not comply
with the distribution of the factual samples in the target class to the
point where they are clearly distinguishable from the ground-truth data.
In panel (b), we have trained our model with CT, once again without any
mutability constraints. We observe that the counterfactuals are highly
plausible, meeting the first objective of Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used CT and protected the immutable feature as described in
Section~\ref{sec-constraints}. We observe that due to the classifier's
lower sensitivity to \emph{age}, recourse recommendations with respect
to \emph{debt} are much more homogenous and do not disproportionately
punish younger individuals. The counterfactuals are also plausible with
respect to the mutable feature and shorter on average than in (c). Thus,
we consider the model in panel (d) as the most explainable according to
Def. \ref{def-explainability}.

@@

In this section, we present experiments that we have conducted in order
to answer the following research questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does our proposed counterfactual training objective in
  Equation~\ref{eq-obj} induce models to learn plausible explanations?
\item
  To what extent does our proposed counterfactual training objective in
  Equation~\ref{eq-obj} yield more favorable algorithmic recourse
  outcomes in the presence of actionability constraints?
\item
  What are the effects of hyperparameter selection with respect to
  Equation~\ref{eq-obj}?
\end{enumerate}

@@

\subsubsection{Evaluation}\label{evaluation}

@@

\subsubsection{Plausibility}\label{sec-plaus}

Table~\ref{tbl-main} presents our main empirical findings. The top five
rows show the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. The following row shows the reduction
in implausibility as measured by Equation~\ref{eq-impl-div} and
aggregated across all test specifications of \emph{ECCo}. The following
two rows show the test accuracies for the model trained with CT and
conventionally trained models (``vanilla'').

We observe that for all datasets except \emph{OL} and across all test
settings, the average distance of counterfactuals from observed samples
in the target class is reduced, indicating improved plausibility. The
magnitude of improvements varies by dataset: for the simple synthetic
datasets, distance reductions range from around 20-40\% (\emph{LS},
\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
datasets, improvements are generally smaller but still substantial in
many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by looking at the absolute outcome. Improvements for other
datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. \emph{MNIST} is the only dataset for which the two metrics
disagree.

@@

\subsubsection{Actionability}\label{sec-act}

@@

\subsubsection{Impact of hyperparameter
settings}\label{sec-hyperparameters}

We test in-depth the impact of three types of hyperparameters; our
complete results are in the appendix.

\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it uses a surrogate VAE to generate
counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
Third, increasing \(T\), the maximum number of steps, generally yields
better outcomes because more CEs can mature in each training epoch.
Fourth, the impact of \(\tau\), the required decision threshold is more
difficult to predict. On ``harder'' datasets it may be difficult to
satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
increasing this threshold does not seem to correlate with better
outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
leads to optimal results because it is associated with high proportions
of mature counterfactuals.

\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not very pronounced.

\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
observed desired improvements in explainability when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT may be a way to
improve the explainability of models in a fine-tuning manner.

@@

We first address the direct extensions of CT in
Section~\ref{sec-approach}. Then, we look at its limitations and
challenges in Section~\ref{sec-limitations}.

\subsection{Future Research}\label{sec-approach}

\textbf{\emph{CT is defined only for classification settings.}} Our
formulation relies on the distinction between non-target class(es)
\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
discrete. Thus, it does not apply to ML tasks where the change in
outcome cannot be readily quantified. Focus on classification models is
a common restriction in research on CEs and AR. Other settings have
attracted some interest (e.g., regression in
\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
little consensus how to robustly extend the notion of counterfactuals.

\textbf{\emph{CT is subject to training instabilities.}} JEMs are
susceptible to instabilities during training \cite{grathwohl2020your}
and even though we depart from the SGLD-based sampling, we still
encounter major variability in the outcomes. CT is exposed to two
potential sources of instabilities: (1) the energy-based contrastive
divergence term in Equation~\ref{eq-div}, and (2) the underlying
counterfactual explainers. Still, we find that training instabilities
can be successfully mitigated by regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each training epoch, and including only mature counterfactuals
for contrastive divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
benefits from the tuning of certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

\subsection{Current Limitations}\label{sec-limitations}

\textbf{\emph{CT increases the training time of models.}} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.

\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature \cite{venkatasubramanian2020philosophical} impacting
the capacity of CT to increase the explainability of the model.

\textbf{\emph{Interventions on features may impact fairness.}} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features, which may have implication for
the fair and equitable treatment of decision subjects. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}. Still,
these phenomena are not specific to CT.

>>>>

\textbf{Specifically, we propose counterfactual training (CT)}: a novel
training regime that aligns learned representations with plausible
explanations respecting actionability constraints. The remainder of this
paper is structured as follows. Section~\ref{sec-lit} presents related
work, focusing on the link between adversarial examples and
counterfactual explanations. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In
Section~\ref{sec-experiments}, through extensive experiments, we
demonstrate that CT substantially improves explainability without
sacrificing predictive performance. We discuss the challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion} that CT is a promising approach towards
making opaque models more trustworthy.

@@

In the natural language processing (NLP) domain, CEs have also been used
to improve models through data augmentation. \cite{wu2021polyjuice}
propose \emph{Polyjuice}, a general-purpose counterfactual generator for
language models. The authors empirically demonstrate that the
augmentation of training data with their method improves robustness in a
number of NLP tasks. \cite{balashankar2023improving} similarly use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure: the
authors identify training samples that are subject to high predictive
uncertainty, generate CEs for those samples, and then fine-tune the
language model on a dataset augmented with the CEs.

@@

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that guarantee
individual recourse to some positive target class with high probability.
Their approach builds on adversarial training by explicitly inducing
susceptibility to targeted adversarial examples for the positive class.
Additionally, the proposed method allows for imposing a set of
actionability constraints ex-ante. For example, users can specify that
certain features are immutable. Secondly, \cite{guo2023counternet} are
the first to propose an end-to-end training pipeline that includes CEs
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to \cite{guo2023counternet},
we impose no restrictions on the ANN architecture at all.

@@

As observed by \cite{grathwohl2020your}, any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

@@

From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this simply in terms ignoring
implausibility costs with respect to immutable features, which
effectively forces the model to instead seek plausibility with respect
to the remaining features. This in turn results in lower overall
sensitivity to immutable features, which we demonstrate empirically for
different classifiers in Section~\ref{sec-experiments}. Under certain
conditions, this result holds theoretically:\footnote{For the proof, see
  the supplementary appendix.}

@@

It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}}
assumes independence of features. This raises a valid concern about the
effect of protecting immutable features in the presence of proxies that
remain unprotected. We address this in Section~\ref{sec-discussion}.

@@

We have simulated this scenario using synthetic data with two
independent features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels show the outcomes for different
training procedures using the same model architectures (a linear
classifier). In panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we used CT. Only in panels
(c) and (d) do we impose the mutability constraint on \emph{age} at test
time. In each case, we show the decision boundary (in green) and the
training data colored according to their ground-truth label: orange
points belong to the target class, \(y^+=2\), blue points belong to the
non-target class, \(y^-=1\). Stars indicate CEs in the target class
generated at test time using generic gradient descent until convergence.

In all cases the counterfactuals are valid but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly distinguishable cluster. In panel (b), they are highly
plausible, meeting the first objective of Def. \ref{def-explainability}.
In panel (c), the CEs involve substantial reductions in \emph{debt} for
younger applicants. By comparison, counterfactual paths are shorter on
average in panel (d) where we have protected the immutable \emph{age} as
described in Section~\ref{sec-constraints}. Due to the classifier's
lower sensitivity to \emph{age}, recommendations with respect to
\emph{debt} are much more homogenous and do not unfairly punish younger
individuals. These counterfactuals are also plausible with respect to
the mutable feature. Thus, we consider the model in panel (d) as the
most explainable according to Def. \ref{def-explainability}.

@@

Here we present the experiments conducted to answer our research
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does our counterfactual training objective as defined
  in Equation~\ref{eq-obj} induce models to learn plausible
  explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

@@

@@

\subsubsection{Plausibility.}\label{sec-plaus}

Table~\ref{tbl-main} presents our main empirical findings. The top five
rows show the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. Row 6 shows the reduction in
implausibility as measured by Equation~\ref{eq-impl-div} and aggregated
across all test specifications of \emph{ECCo}. Rows 7 and 8 show the
test accuracies for the model trained with CT and conventionally trained
models (``vanilla'').

For all datasets except \emph{OL} and across all test settings, the
average distance of CEs from observed samples in the target class is
reduced, indicating improved plausibility. The magnitude of improvements
varies. For the simple synthetic datasets, distance reductions range
from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
(\emph{Circ}). For the real-world tabular datasets, improvements tend to
be smaller but still substantial, with around 10-15\% for \emph{CH},
11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred} and around 3\% for
\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
reduced by up to 9\%. The results for our proposed divergence metric are
qualitatively similar, but generally even more pronounced: for the
\emph{Circ} dataset, implausibility is reduced by almost 94\% to
virtually zero as we verified by the absolute outcome. Improvements for
other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. \emph{MNIST} is the only dataset for which the two metrics
disagree.

@@

\subsubsection{Actionability.}\label{sec-act}

@@

\subsubsection{Impact of hyperparameter
settings.}\label{sec-hyperparameters}

We test the impact of three types of hyperparameters; our complete
results are in the supplementary appendix.\\
\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it uses a surrogate VAE to generate
counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
Third, increasing \(T\), the maximum number of steps, generally yields
better outcomes because more CEs can mature in each training epoch.
Fourth, the impact of \(\tau\), the required decision threshold is more
difficult to predict. On ``harder'' datasets it may be difficult to
satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
increasing this threshold does not seem to correlate with better
outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
leads to optimal results because it is associated with high proportions
of mature counterfactuals.\\
\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
observed desired improvements in explainability when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT may be a way to
improve the explainability of models in a fine-tuning manner.

@@

As our results indicate, counterfactual training produces models that
are more explainable. However, our approach is not without
limitations.\\
\textbf{\emph{CT increases the training time of models.}} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature \cite{venkatasubramanian2020philosophical} impacting
the capacity of CT to fulfill its intended goal.\\
\textbf{\emph{Interventions on features may impact fairness.}} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features, which may have implication for
the fair and equitable treatment of decision subjects. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}. Still,
these phenomena are not specific to CT.

We also highlight several interesting directions for future research.\\
\textbf{\emph{Extending CT beyond classification settings.}} Our
formulation relies on the distinction between non-target class(es)
\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
discrete. Thus, it does not apply to ML tasks where the change in
outcome cannot be readily quantified. Focus on classification models is
a common restriction in research on CEs and AR. Other settings have
attracted some interest (e.g., regression in
\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
little consensus how to robustly extend the notion of CEs.\\
\textbf{\emph{Addressing the training instabilities.}} JEMs are
susceptible to instabilities during training \cite{grathwohl2020your}
and even though we depart from the SGLD-based sampling, we still
encounter major variability in the outcomes. CT is exposed to two
potential sources of instabilities: (1) the energy-based contrastive
divergence term in Equation~\ref{eq-div}, and (2) the underlying
counterfactual explainers. Still, we find that training instabilities
can be successfully mitigated by regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each training epoch, and including only mature counterfactuals
for contrastive divergence.\\
\textbf{\emph{Improving hyperparameter selection procedures.}} Our
method benefits from the tuning of certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

### 2025-03-14 paper/sections/discussion.qmd
Shorten to slightly less than 16 pages by restructuring

We first address the direct extensions of CT in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.

## Future Research {#sec-approach}

***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of counterfactuals.

***CT is subject to training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.

***CT is sensitive to hyperparameter selection.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

## Current Limitations {#sec-limitations}

***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.

***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.

***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.

>>>>

As our results indicate, counterfactual training produces models that are more explainable. However, our approach is not without limitations.  
***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.

We also highlight several interesting directions for future research.  
***Extending CT beyond classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
***Addressing the training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.  
***Improving hyperparameter selection procedures.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
